---
title: "tag-reconstructor"
author: "Fuad Ale-Enriquez"
date: "2025-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
set.seed (150877)
```

# Overview

This project implements a bottom-up optimization framework to reconstruct a combinatorial set of triacylglyceride (TAG) molecules that best reproduces an experimental fatty acid methyl ester (FAME) profile.

Starting from measured FAME proportions, the method identifies the optimal combination and relative abundances of TAGs whose collective fatty acid composition matches the observed data.

## Required Input

The pipeline requires a tidy dataset containing fatty acid (FAME) proportions with, at minimum, the following columns:

- `Treatment_ID` — experimental condition or sample label  
- `Type` — sample grouping (e.g., treatment class)  
- `FAME` — fatty acid identifier (e.g., C18:1)  
- `Mean_abundance` — abundance or proportion of each fatty acid  
- `rank` — optional ranking of fatty acids by abundance  

## Workflow

The analysis consists of four main steps:

1. **TAG combinatorial generation**  
   Construct all candidate TAG molecules from the observed fatty acid pool.

2. **Optimization of TAG mixtures**  
   Identify a minimal set of TAGs and their proportions that best approximate the experimental FAME profile.

3. **Mass-based conversion to FAME composition**  
   Translate TAG mixtures into predicted FAME distributions using molecular mass relationships.

4. **Comparison with experimental data**  
   Evaluate model performance using quantitative metrics and visual diagnostics against GC-FID measurements.
   
This approach provides a mechanistic link between TAG composition and measurable fatty acid profiles, enabling interpretation and simulation of lipid systems.


# Libraries
```{r}
library(gtools)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(tibble)
library(ggplot2)
library(scales)
library(cowplot)
library(openxlsx)

pkgs <- c("gtools","dplyr","tidyr","stringr","purrr","tibble","ggplot2","scales","cowplot","openxlsx")
missing <- pkgs[!sapply(pkgs, requireNamespace, quietly = TRUE)]
if (length(missing)) stop("Missing packages: ", paste(missing, collapse = ", "))
```

# Functions

## 1. Build experimental composition per (Treatment_ID, Type)

**Default top_n FAME = 12**

**normalize = FALSE is default. Change to TRUE only if Mean_abundance is given as concentration (absolute) and the sum within each sample is not equal to 1**

**use_rank = TRUE is default**

```{r}
get_exp_composition_for_group <- function(top_by_group, treatment_id, type,
                                         top_n = 12, normalize = FALSE, use_rank = TRUE) {
  stopifnot(all(c("Treatment_ID","Type","FAME","Mean_abundance","rank") %in% names(top_by_group)))
  df <- subset(top_by_group, Treatment_ID == treatment_id & Type == type)
  if (nrow(df) == 0) stop(sprintf("No rows for Treatment_ID=%s, Type=%s", treatment_id, type))

  # Select top_n fatty acids
  if (use_rank) {
    df <- df[order(df$rank), ]
  } else {
    df <- df[order(df$Mean_abundance, decreasing = TRUE), ]
  }
  df <- head(df, top_n)

  df$FAME <- as.character(df$FAME)
  df$Mean_abundance <- as.numeric(df$Mean_abundance)

  comp <- tapply(df$Mean_abundance, df$FAME, sum)
  comp[is.na(comp)] <- 0

  if (normalize) {
    s <- sum(comp)
    if (s <= 0) stop("Sum of Mean_abundance for selected FAs is zero.")
    comp <- comp / s
  }

  list(fatty_acids = names(comp), exp_composition = comp)
}
```


## 2. Generate TAG candidates (3 FA per TAG), allowing FA repetition. Collapses stereoisomers by sorting within TAG

```{r}
generate_TAGs_with_repeat <- function(fatty_acids) {
  cmb <- gtools::combinations(length(fatty_acids), 3, v = fatty_acids, repeats.allowed = TRUE)
  TAGs <- unique(t(apply(cmb, 1, function(row) sort(as.character(row)))))
  storage.mode(TAGs) <- "character"
  TAGs
}

generate_TAGs_no_repeat <- function(fatty_acids) {
  TAGs <- t(combn(fatty_acids, 3, simplify = TRUE))
  storage.mode(TAGs) <- "character"
  TAGs
}
```


## 3. Convert a TAG into a composition vector over the FA universe (fractions over that TAG’s 3 positions; sums to 1 across FAs present in the universe)
```{r}
tag_comp <- function(tag_vec, fa_univ) {
  tg <- as.character(tag_vec)
  composition <- table(tg) / length(tg)  # divide by 3
  full <- setNames(rep(0, length(fa_univ)), fa_univ)
  common <- intersect(names(composition), fa_univ)
  if (length(common) > 0) full[common] <- as.numeric(composition[common])
  full
}
```


## 4. Objective for K-TAG mixture with scale handling and FA weights
```{r}
objective_Ktag <- function(x, C_list, exp_comp,
                           metric = "L1", fa_w = NULL,
                           scale_mode = c("match_sum","free")) {
  scale_mode <- match.arg(scale_mode)
  exp_names <- names(exp_comp)
  P <- length(exp_comp)

  # Align C_list to exp_comp names/order
  C_list_aligned <- lapply(C_list, function(v) {
    if (identical(names(v), exp_names)) return(v)
    vv <- setNames(rep(0, P), exp_names)
    common <- intersect(names(v), exp_names)
    if (length(common) > 0) vv[common] <- v[common]
    vv
  })

  Cmat <- do.call(rbind, C_list_aligned)

  K <- length(C_list_aligned)
  logits <- c(if (K > 1) x[1:(K-1)] else numeric(0), 0)
  p <- exp(logits - max(logits)); p <- p / sum(p)
  dim(p) <- c(1L, length(p))

  pred_frac <- as.vector(p %*% Cmat)

  if (scale_mode == "match_sum") {
    S <- sum(exp_comp)
    pred <- S * pred_frac
  } else {
    log_s <- if (length(x) >= K) x[K] else 0
    s <- exp(log_s)
    pred <- s * pred_frac
  }

  if (!is.null(fa_w)) {
    fa_w <- fa_w[exp_names]
    fa_w[is.na(fa_w)] <- 1
  }

  if (metric == "SSE") {
    res <- (pred - exp_comp)^2
    if (!is.null(fa_w)) res <- fa_w * res
    sum(res)
  } else {
    res <- abs(pred - exp_comp)
    if (!is.null(fa_w)) res <- fa_w * res
    sum(res)
  }
}
```

## 5. Optimize weights for a given TAG set; multi-start
```{r}
optimize_one_Ktag_combo <- function(TAG_list, fa_univ, exp_comp,
                                   metric = "L1", fa_w = NULL,
                                   scale_mode = "match_sum",
                                   n_starts = 5, start_scale = 1.0,
                                   brent_logit_bounds = c(-20, 20)) {
  K <- length(TAG_list)
  C_list <- lapply(TAG_list, function(tg) tag_comp(tg, fa_univ))

  # K=1 closed form
  if (K == 1) {
    pred_frac <- C_list[[1]]
    s <- sum(exp_comp)
    pred <- s * pred_frac

    fa_w_loc <- if (!is.null(fa_w)) { w <- fa_w[names(exp_comp)]; w[is.na(w)] <- 1; w } else NULL

    if (metric == "SSE") {
      res <- (pred - exp_comp)^2
      if (!is.null(fa_w_loc)) res <- fa_w_loc * res
      err <- sum(res)
    } else {
      res <- abs(pred - exp_comp)
      if (!is.null(fa_w_loc)) res <- fa_w_loc * res
      err <- sum(res)
    }

    return(list(weights = 1, scale = s, error = err, convergence = 0))
  }

  dim_logits <- K - 1
  add_scale <- (scale_mode == "free")

  # Special case: K=2, match_sum => 1D Brent
  if (dim_logits == 1 && !add_scale) {
    oneD_obj <- function(t) {
      objective_Ktag(x = c(t), C_list = C_list, exp_comp = exp_comp,
                     metric = metric, fa_w = fa_w, scale_mode = scale_mode)
    }
    opt_res <- optimize(oneD_obj, interval = brent_logit_bounds)
    logits <- c(opt_res$minimum, 0)
    p <- exp(logits - max(logits)); p <- p / sum(p)
    return(list(weights = p, scale = sum(exp_comp), error = opt_res$objective, convergence = 0))
  }

  # Multi-start Nelder–Mead
  build_starts <- function(dim_logits, add_scale) {
    base <- diag(dim_logits) * 0.5
    base <- rbind(rep(0, dim_logits), base)
    if (n_starts > nrow(base)) {
      extra <- matrix(rnorm((n_starts - nrow(base)) * dim_logits, sd = start_scale),
                      ncol = dim_logits)
      starts <- rbind(base, extra)
    } else {
      starts <- base[seq_len(n_starts), , drop = FALSE]
    }
    if (add_scale) {
      S0 <- sum(exp_comp)
      starts <- cbind(starts, rep(log(max(S0, .Machine$double.eps)), nrow(starts)))
    }
    starts
  }

  starts <- build_starts(dim_logits, add_scale)
  best <- list(value = Inf, par = NULL, conv = NA_integer_)

  for (i in seq_len(nrow(starts))) {
    res <- optim(
      par = starts[i, ],
      fn  = objective_Ktag,
      C_list = C_list,
      exp_comp = exp_comp,
      metric = metric,
      fa_w = fa_w,
      scale_mode = scale_mode,
      method = "Nelder-Mead",
      control = list(maxit = 6000)
    )
    if (res$value < best$value) best <- list(value = res$value, par = res$par, conv = res$convergence)
  }

  logits <- c(best$par[1:dim_logits], 0)
  p <- exp(logits - max(logits)); p <- p / sum(p)

  s <- if (scale_mode == "match_sum") {
    sum(exp_comp)
  } else {
    exp(best$par[dim_logits + 1])
  }

  list(weights = p, scale = s, error = best$value, convergence = best$conv)
}
```


## 6. Candidate ranking and greedy forward selection
```{r}
rank_TAG_candidates <- function(TAGs, fatty_acids, exp_comp) {
  exp_frac <- exp_comp / sum(exp_comp)
  scores <- apply(TAGs, 1, function(tag) {
    cvec <- tag_comp(tag, fatty_acids)
    num <- sum(cvec * exp_frac)
    den <- sqrt(sum(cvec^2)) * sqrt(sum(exp_frac^2))
    if (den == 0) 1 else (1 - num / den)
  })
  ord <- order(scores)
  list(order = ord, scores = scores)
}

run_K_TAG_search_greedy <- function(fatty_acids, exp_comp,
                                   K = 5,
                                   metric = "L1", fa_w = NULL,
                                   scale_mode = "match_sum",
                                   n_starts = 5,
                                   allow_within_TAG_repeat = TRUE,
                                   candidate_limit = 120) {

  TAGs <- if (allow_within_TAG_repeat) {
    generate_TAGs_with_repeat(fatty_acids)
  } else {
    generate_TAGs_no_repeat(fatty_acids)
  }
  if (nrow(TAGs) == 0) stop("No TAG candidates.")

  rk <- rank_TAG_candidates(TAGs, fatty_acids, exp_comp)
  ord <- rk$order
  keep_n <- min(candidate_limit, length(ord))
  pool <- TAGs[ord[seq_len(keep_n)], , drop = FALSE]

  selected <- list()
  selected_rows <- integer(0)
  best_opt <- NULL

  for (t in 1:K) {
    best_t <- list(err = Inf, row = NA, opt = NULL)

    for (r in seq_len(nrow(pool))) {
      if (r %in% selected_rows) next
      trial_list <- c(selected, list(pool[r, ]))
      opt <- optimize_one_Ktag_combo(
        TAG_list = trial_list,
        fa_univ = fatty_acids,
        exp_comp = exp_comp,
        metric = metric, fa_w = fa_w,
        scale_mode = scale_mode,
        n_starts = n_starts
      )
      if (opt$error < best_t$err) best_t <- list(err = opt$error, row = r, opt = opt)
    }

    selected <- c(selected, list(pool[best_t$row, ]))
    selected_rows <- c(selected_rows, best_t$row)
    best_opt <- best_t$opt
  }

  sel_TAGs <- selected
  names(sel_TAGs) <- paste0("T", seq_len(K))
  w <- best_opt$weights
  scl <- best_opt$scale

  C_list <- lapply(sel_TAGs, function(tg) tag_comp(tg, fatty_acids))
  Cmat <- do.call(rbind, C_list)
  pred_abs <- as.numeric(scl * (w %*% Cmat))
  names(pred_abs) <- fatty_acids

  list(
    TAGs = sel_TAGs,
    weights = w,
    scale = scl,
    error = best_opt$error,
    convergence = best_opt$convergence,
    predicted_absolute = pred_abs
  )
}
```


## 7. Run per (Treatment_ID, Type) across data
```{r}
run_all_groups_5TAG <- function(top_by_group,
                               top_n = 12, normalize = FALSE, use_rank = TRUE,
                               metric = "L1", fa_w = NULL,
                               scale_mode = "match_sum",
                               n_starts = 5,
                               allow_within_TAG_repeat = TRUE,
                               candidate_limit = 120) {

  groups <- unique(top_by_group[, c("Treatment_ID","Type")])
  groups$Treatment_ID <- as.character(groups$Treatment_ID)
  groups$Type <- as.character(groups$Type)

  out <- vector("list", nrow(groups))
  predicted_list <- vector("list", nrow(groups))
  residuals_list <- vector("list", nrow(groups))

  for (g in seq_len(nrow(groups))) {
    trt <- groups$Treatment_ID[g]
    typ <- groups$Type[g]

    exp_info <- get_exp_composition_for_group(
      top_by_group, trt, typ,
      top_n = top_n, normalize = normalize, use_rank = use_rank
    )

    fatty_acids <- exp_info$fatty_acids
    exp_comp <- exp_info$exp_composition[fatty_acids]
    total_exp <- sum(exp_comp)
    FA_count <- length(fatty_acids)

    fit <- run_K_TAG_search_greedy(
      fatty_acids = fatty_acids,
      exp_comp = exp_comp,
      K = 5,
      metric = metric,
      fa_w = fa_w,
      scale_mode = scale_mode,
      n_starts = n_starts,
      allow_within_TAG_repeat = allow_within_TAG_repeat,
      candidate_limit = candidate_limit
    )

    TAG_strings <- vapply(fit$TAGs, function(tg) paste(as.character(tg), collapse = "-"), character(1))
    ws <- fit$weights
    resid <- fit$predicted_absolute - exp_comp
    avg_error_perFA <- if (FA_count > 0) fit$error / FA_count else NA_real_

    out[[g]] <- data.frame(
      Treatment_ID = trt,
      Type = typ,
      FA_count = FA_count,
      exp_sum = total_exp,
      TAG1 = TAG_strings[1],
      TAG2 = TAG_strings[2],
      TAG3 = TAG_strings[3],
      TAG4 = TAG_strings[4],
      TAG5 = TAG_strings[5],
      w1 = ws[1], w2 = ws[2], w3 = ws[3], w4 = ws[4], w5 = ws[5],
      scale = fit$scale,
      error = fit$error,
      avg_error_perFA = avg_error_perFA,
      convergence = fit$convergence,
      stringsAsFactors = FALSE
    )

    predicted_list[[g]] <- tibble(
      Treatment_ID = trt, Type = typ,
      FAME = names(fit$predicted_absolute),
      predicted = as.numeric(fit$predicted_absolute)
    )

    residuals_list[[g]] <- tibble(
      Treatment_ID = trt, Type = typ,
      FAME = names(resid),
      residual = as.numeric(resid)
    )
  }

  res <- do.call(rbind, out)

  # Attach detail tables safely to the returned object
  attr(res, "predicted_list") <- bind_rows(predicted_list)
  attr(res, "residuals_list") <- bind_rows(residuals_list)

  res
}
```


# USAGE EXAMPLE - Computation (GitHub-friendly)

## IMPORT
```{r}
# =========================
# Data loader: use object if present, else read from repo file
# If running on R Studio environment it is better to import top12_by_group and FAMEprofile prior running this chunk
# =========================

load_or_read <- function(object_name, file_path, required = TRUE) {
  if (exists(object_name, envir = .GlobalEnv)) {
    message("Using existing object: ", object_name)
    return(get(object_name, envir = .GlobalEnv))
  }
  if (file.exists(file_path)) {
    message("Reading from file: ", file_path)
    return(read.csv(file_path, stringsAsFactors = FALSE))
  }
  if (required) {
    stop("Missing required input: '", object_name, "'. ",
         "Provide it in the environment or add file: ", file_path)
  } else {
    message("Optional input not found: ", object_name, 
            ". Skipping sections that require it.")
    return(NULL)
  }
}

# Required input for fitting
top_by_group <- load_or_read("top12_by_group", file.path("data", "example_top12_by_group.csv"), required = TRUE)

# Optional input for triplicate-vs-sim evaluation plots
FAMEprofile <- load_or_read("FAMEprofile", file.path("data", "example_FAMEprofile.csv"), required = FALSE)
```


## SIMULATION
```{r}
#Run the pipeline
set.seed(150877)  # for reproducibility of starts

results_5TAG <- run_all_groups_5TAG(
  top_by_group = top_by_group,  # your existing data frame
  top_n = 12,
  normalize = FALSE,        
  use_rank = TRUE,
  metric = "SSE",            
  fa_w = NULL,              
  scale_mode = "match_sum", 
  n_starts = 8,
  allow_within_TAG_repeat = TRUE,
  candidate_limit = 200
)

print(head(results_5TAG))

# Access extra outputs (if you used the attr() fix)
pred_list  <- attr(results_5TAG, "predicted_list")
resid_list <- attr(results_5TAG, "residuals_list")
```

## EXPORTING SIMULATED TAG RESULTS
```{r}
dir.create("outputs", showWarnings = FALSE)
dir.create(file.path("outputs", "tables"), showWarnings = FALSE, recursive = TRUE)

# If you used the improved run_all_groups_5TAG() that stores these as attributes:
pred_list  <- attr(results_5TAG, "predicted_list")
resid_list <- attr(results_5TAG, "residuals_list")

wb <- createWorkbook()

addWorksheet(wb, "summary")
writeData(wb, "summary", results_5TAG)

if (!is.null(pred_list)) {
  addWorksheet(wb, "predicted")
  writeData(wb, "predicted", pred_list)
}

if (!is.null(resid_list)) {
  addWorksheet(wb, "residuals")
  writeData(wb, "residuals", resid_list)
}

saveWorkbook(
  wb,
  file = file.path("outputs", "tables", "results_5TAG_with_extra.xlsx"),
  overwrite = TRUE
)
```


## SIMULATION FITNESS EXAMPLE - Three technical replicates vs. simulated example - ice cream waste paper
**Requires pipeline output "results_5TAG" + compatible unsummarized FAME profile "FAMEprofile"**

### Computing simulation fitness
```{r}
stopifnot(exists("results_5TAG"))

if (is.null(FAMEprofile)) {
  message("FAMEprofile not available. Skipping triplicate-vs-simulated fitness section.")
  knitr::knit_exit()
}

theme_set(theme_minimal(base_size = 12))

# Color-blind safe palette
col_exp <- "#0072B2"  # experimental dots
col_sim <- "#E69F00"  # simulated diamonds

# Controls
THRESH   <- 1.1        # filter experimental points below this % (w/w)
ORDER_BY <- "exp_mean" # exp_mean or sim_mean

# -------------------------
# Label utilities (define ONCE)
# -------------------------
sanitize_label <- function(x) {
  x %>%
    str_replace_all("[\\u00A0\\u2007\\u202F\\u200B\\uFEFF]", " ") %>%
    str_replace_all("[\\u2010\\u2011\\u2012\\u2013\\u2014\\u2212]", "-") %>%
    str_replace_all("\\s+", " ") %>%
    str_replace_all("\\(\\s+", "(") %>%
    str_replace_all("\\s+\\)", ")") %>%
    str_trim()
}

repair_label <- function(x) {
  x %>%
    str_replace("\\bUnknown\\s*3\\b", "C18:1 (E9)") %>%
    str_replace("C22:2\\s*\\?\\?\\?", "C22:2 (Z13,16)")
}

repair_sim_tag <- function(tag_vec) {
  tag_vec %>%
    str_squish() %>%
    str_replace("\\bUnknown\\s*3\\b", "C18:1 (E9)") %>%
    str_replace("C22:2\\s*\\?\\?\\?", "C22:2 (Z13,16)")
}

# -------------------------
# Chemistry helpers
# -------------------------
amu <- list(C = 12.011, H = 1.00794, O = 15.999)

mw_fame <- function(n, d) {
  C <- n + 1
  H <- 2*n - 2*d + 2
  O <- 2
  C*amu$C + H*amu$H + O*amu$O
}

mw_tag <- function(n_vec, d_vec) {
  sc <- sum(n_vec); sd <- sum(d_vec)
  C <- 3 + sc
  H <- 2 + 2*sc - 2*sd
  O <- 6
  C*amu$C + H*amu$H + O*amu$O
}

parse_chain <- function(x) {
  x1 <- sanitize_label(x) %>% repair_label()
  m  <- str_match(x1, "C(\\d+):(\\d+)")
  if (any(is.na(m))) stop("Could not parse chain: ", x)
  tibble(label = x1, n = as.integer(m[,2]), d = as.integer(m[,3]))
}

parse_tag <- function(tag_str) {
  parts <- str_split(tag_str, "-", simplify = TRUE) %>% as.vector()
  bind_rows(map(parts, parse_chain))
}

# -------------------------
# Simulated (SAE): mass-correct %FAME from 1.000 g TAG mixture
# -------------------------
sim_fame_masscorrect <- results_5TAG %>%
  filter(Type == "SAE") %>%
  mutate(across(matches("^TAG\\d$"), ~ repair_sim_tag(.x))) %>%
  pivot_longer(
    cols = matches("^TAG\\d$|^w\\d$"),
    names_to = c(".value","idx"),
    names_pattern = "([A-Za-z]+)(\\d+)"
  ) %>%
  mutate(TAG = str_squish(TAG)) %>%
  filter(!is.na(TAG), !is.na(w)) %>%
  group_by(Treatment_ID) %>%
  group_modify(~{
    w_frac <- .x$w / sum(.x$w, na.rm = TRUE)  # normalize weights
    .x2 <- mutate(.x, w_g = w_frac)          # grams TAG in 1.000 g mixture

    contrib <- imap_dfr(.x2$TAG, function(tag_str, i) {
      w_g  <- .x2$w_g[i]
      ch   <- parse_tag(tag_str)
      Mtag <- mw_tag(ch$n, ch$d)
      nTAG <- if (Mtag > 0) w_g / Mtag else 0

      ch %>%
        mutate(label = sanitize_label(repair_label(label)),
               mass_g = nTAG * mw_fame(n, d)) %>%
        select(label, mass_g)
    })

    total_fame_g <- sum(contrib$mass_g, na.rm = TRUE)

    contrib %>%
      group_by(label) %>%
      summarise(fame_g = sum(mass_g, na.rm = TRUE), .groups = "drop") %>%
      mutate(`%FAME_sim_w` = 100 * fame_g / total_fame_g)
  }) %>%
  ungroup() %>%
  group_by(Treatment_ID, label) %>%
  summarise(`%FAME_sim_w` = sum(`%FAME_sim_w`), .groups = "drop") %>%
  mutate(label = sanitize_label(repair_label(label)))

# -------------------------
# Experimental (SAE): %FAME as percent, filtered at THRESH
# -------------------------
exp_by_rep <- FAMEprofile %>%
  filter(Type == "SAE") %>%
  transmute(
    Treatment_ID,
    label_raw = sanitize_label(repair_label(`Peak#`)),
    `%FAME_exp_w` = `%FAME` * 100,
    Rep_ID = Lab_Tag
  ) %>%
  filter(`%FAME_exp_w` >= THRESH)

# -------------------------
# Shared axis ordering (global levels)
# -------------------------
exp_mean_global <- exp_by_rep %>%
  group_by(label_raw) %>%
  summarise(exp_mean = mean(`%FAME_exp_w`, na.rm = TRUE), .groups = "drop") %>%
  transmute(label = sanitize_label(label_raw), exp_mean)

sim_global <- sim_fame_masscorrect %>%
  group_by(label) %>%
  summarise(sim_mean = mean(`%FAME_sim_w`, na.rm = TRUE), .groups = "drop")

all_labels <- union(sim_global$label, exp_mean_global$label) %>% sanitize_label()

ord_df <- tibble(label = all_labels) %>%
  left_join(exp_mean_global, by = "label") %>%
  left_join(sim_global, by = "label") %>%
  mutate(order_val = if (ORDER_BY == "exp_mean") ifelse(is.na(exp_mean), Inf, exp_mean)
                     else ifelse(is.na(sim_mean), Inf, sim_mean)) %>%
  arrange(order_val, label)

levels_global <- unique(ord_df$label[nzchar(ord_df$label)])

# Factorize for plotting
sim_plot <- sim_fame_masscorrect %>%
  mutate(FAME = factor(sanitize_label(label), levels = levels_global))

exp_plot <- exp_by_rep %>%
  mutate(FAME = factor(sanitize_label(label_raw), levels = levels_global))
```

### Plot example
```{r}
gg_rep_vs_sim <- ggplot() +
  geom_point(
    data = sim_plot,
    aes(x = FAME, y = `%FAME_sim_w`, shape = "Simulated"),
    size = 3.0, stroke = 0.3,
    color = "grey20", fill = col_sim
  ) +
  geom_point(
    data = exp_plot,
    aes(x = FAME, y = `%FAME_exp_w`, shape = "Experimental"),
    position = position_jitter(width = 0.10, height = 0, seed = 1),
    size = 1.8, alpha = 0.65,
    color = col_exp
  ) +
  coord_flip() +
  scale_x_discrete(drop = FALSE) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  scale_shape_manual(
    name   = NULL,
    breaks = c("Experimental", "Simulated"),
    labels = c("Experimental (GC-FID)", "Simulated"),
    values = c("Experimental" = 16, "Simulated" = 23)
  ) +
  guides(
    shape = guide_legend(
      override.aes = list(
        color = c(col_exp, "grey20"),
        fill  = c(NA, col_sim),
        size  = c(1.8, 3.0),
        alpha = 1
      )
    )
  ) +
  labs(x = "Fatty acid residue", y = "Percent of total FAME (w/w)") +
  facet_wrap(
    ~Treatment_ID,
    ncol = 2, scales = "free_y",
    labeller = labeller(
      Treatment_ID = c(
        "High fat" = "High initial ICW fat content (8.0%w)",
        "Low fat"  = "Low initial ICW fat content (4.5%w)"
      )
    )
  ) +
  theme(
    legend.position = "top",
    axis.text = element_text(color = "black"),
    axis.text.y = element_text(size = 9),
    panel.grid = element_blank(),
    panel.grid.major.y = element_line(colour = alpha("grey40", 0.18), linewidth = 0.3),
    axis.line = element_line(color = "black", linewidth = 0.45),
    axis.ticks = element_line(color = "black", linewidth = 0.45),
    strip.text = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey92", color = NA)
  )

gg_rep_vs_sim
```

### Fitness and diagnostic metrics
```{r}
# -------------------------
# Align on shared axis per treatment
# -------------------------
grid_treat <- expand_grid(
  Treatment_ID = sort(unique(c(sim_fame_masscorrect$Treatment_ID, exp_by_rep$Treatment_ID))),
  FAME = factor(levels_global, levels = levels_global)
)

sim_mat_treat <- sim_fame_masscorrect %>%
  mutate(FAME = factor(sanitize_label(label), levels = levels_global)) %>%
  select(Treatment_ID, FAME, sim = `%FAME_sim_w`) %>%
  right_join(grid_treat, by = c("Treatment_ID","FAME")) %>%
  mutate(sim = coalesce(sim, 0))

exp_mean_treat <- exp_by_rep %>%
  mutate(FAME = factor(sanitize_label(label_raw), levels = levels_global)) %>%
  group_by(Treatment_ID, FAME) %>%
  summarise(exp = mean(`%FAME_exp_w`, na.rm = TRUE), .groups = "drop") %>%
  right_join(grid_treat, by = c("Treatment_ID","FAME")) %>%
  mutate(exp = coalesce(exp, 0))

fit_treat_df <- sim_mat_treat %>%
  left_join(exp_mean_treat, by = c("Treatment_ID","FAME"))

metrics_from_vectors <- function(sim, exp) {
  sim <- as.numeric(sim); exp <- as.numeric(exp)
  rmse <- sqrt(mean((sim - exp)^2))
  mae  <- mean(abs(sim - exp))
  bias <- mean(sim - exp)
  r <- if (sd(sim) > 0 && sd(exp) > 0) suppressWarnings(cor(sim, exp)) else NA_real_
  denom <- sqrt(sum(sim^2)) * sqrt(sum(exp^2))
  cosine <- if (denom > 0) sum(sim * exp) / denom else NA_real_
  bc <- if (sum(sim + exp) > 0) sum(abs(sim - exp)) / sum(sim + exp) else NA_real_
  tibble(RMSE = rmse, MAE = mae, Bias = bias, Pearson_r = r, Cosine = cosine, BrayCurtis = bc)
}

fitness_treatment <- fit_treat_df %>%
  group_by(Treatment_ID) %>%
  summarise(metrics_from_vectors(sim, exp), .groups = "drop")

fitness_treatment

facet_labeller <- labeller(
  Treatment_ID = c(
    "High fat" = "High initial ICW fat content (8.0%w)",
    "Low fat"  = "Low initial ICW fat content (4.5%w)"
  )
)

# Scatter
scatter_df <- fit_treat_df %>% mutate(Exp_mean = exp, Sim = sim)

gg_fitness_scatter <- ggplot(scatter_df, aes(x = Exp_mean, y = Sim)) +
  geom_abline(slope = 1, intercept = 0, color = "grey60", linetype = "dashed") +
  geom_point(size = 2.0, alpha = 0.85, color = "darkred") +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6) +
  scale_x_continuous(limits = c(0, 30), breaks = seq(0, 30, by = 5), labels = percent_format(scale = 1)) +
  scale_y_continuous(limits = c(0, 30), breaks = seq(0, 30, by = 5), labels = percent_format(scale = 1)) +
  labs(x = "Experimental percent of total FAME (w/w)",
       y = "Simulated percent of total FAME (w/w)") +
  facet_wrap(~Treatment_ID, ncol = 2, scales = "free_y", labeller = facet_labeller) +
  theme(
    panel.grid = element_blank(),
    axis.line  = element_line(color = "black", linewidth = 0.45),
    axis.ticks = element_line(color = "black", linewidth = 0.45),
    axis.text  = element_text(color = "black"),
    strip.text = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey92", color = NA)
  )

# Residual bars
resid_df <- fit_treat_df %>%
  mutate(Residual = sim - exp)

gg_residual_bars <- ggplot(resid_df, aes(x = FAME, y = Residual, fill = Residual > 0)) +
  geom_hline(yintercept = 0, color = "grey20", linewidth = 0.4) +
  geom_col(width = 0.65, show.legend = FALSE) +
  coord_flip() +
  scale_x_discrete(drop = FALSE) +
  scale_y_continuous(limits = c(-10, 10), breaks = seq(-10, 10, by = 5), labels = percent_format(scale = 1)) +
  scale_fill_manual(values = c("TRUE" = "#D55E00", "FALSE" = "#0072B2")) +
  labs(x = "Fatty acid residue", y = "Residual (Sim - Exp)") +
  facet_wrap(~Treatment_ID, ncol = 2, scales = "free_y", labeller = facet_labeller) +
  theme(
    panel.grid = element_blank(),
    axis.line  = element_line(color = "black", linewidth = 0.45),
    axis.ticks = element_line(color = "black", linewidth = 0.45),
    axis.text  = element_text(color = "black"),
    strip.text = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey92", color = NA)
  )

# Combine panels (A, C, B like your original)
pA <- gg_rep_vs_sim + theme(plot.margin = margin(6, 10, 6, 12))
pB <- gg_fitness_scatter + theme(plot.margin = margin(6, 10, 6, 12))
pC <- gg_residual_bars + theme(plot.margin = margin(6, 10, 6, 12))

aligned <- cowplot::align_plots(pA, pB, pC, align = "v", axis = "l")
panel_ABC <- cowplot::plot_grid(
  aligned[[1]],
  aligned[[3]],
  aligned[[2]],
  ncol = 1,
  labels = c("A", "B", "C"),
  label_size = 14,
  rel_heights = c(1.4, 1.2, 1.2)
)

panel_ABC

dir.create(file.path("outputs", "figures"), showWarnings = FALSE, recursive = TRUE)

ggsave(
  filename = file.path("outputs", "figures", "Figure3_simulation_vs_experimental.jpeg"),
  plot = panel_ABC,
  width = 7,
  height = 10.5,
  dpi = 300
)

```

